Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed particle hydrodynamics

### 2017-05-07 Manuscripts_GMD.tex
a small edit

While SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM (Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediately dynamics and thermodynamics equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still coarse compared with SK-3D and PDAC.

@@

As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consisting of solid with different size and mixture of gases is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediately thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases does not show up under this assumption). Immediately dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term does not show up with this assumption).

@@

\subsection{Governing equations}
Based on above assumptions, the governing equations of our model are given as following, which is the same as the governing equations of SK-3D \citep{suzuki2005numerical}.
\begin{align}
\dfrac{\partial \rho}{\partial t} + \nabla \cdot \left(\rho \textbf{v}\right) = 0 \label{eq:gov-cs-rho} \\
\dfrac{\partial \rho \xi}{\partial t} + \nabla \cdot \left(\rho \xi \textbf{v}\right) = 0 \label{eq:gov-cs-ks}\\
\dfrac{\partial \rho \textbf{v}}{\partial t} + \nabla \cdot \left(\rho \textbf{v} \textbf{v} + p\textbf{I}\right) = \rho \textbf{g} \label{eq:gov-cs-v} \\
\dfrac{\partial \rho E}{\partial t} + \nabla \cdot \left[\left(\rho E + p \right)\textbf{v}\right] = \rho \textbf{g} \cdot\textbf{v} \label{eq:gov-cs-e}
\end{align}
Where $\rho$ is the density, $\textbf{v}$ is the velocity, $\xi$ is the mass fraction of ejected material, $\textbf{g}$ is the gravitational acceleration, $\textbf{I}$ is a unit tensor.
$E = e + K $ is the total energy which is a summation of kinetic energy $K$ and internal energy $e$.
An additional equation is required to close the system. In this model, the equation for closing the system is the following EOS (equation of state).
\begin{equation}
p = \left(\gamma_m - 1\right)\rho e \label{eq:EOS}
\end{equation}
Where 
\begin{equation}
\gamma_m = R_m/C_{vm} + 1 \label{eq:gov-gm}
\end{equation}
\begin{equation}
R_m = \xi_g R_g + \xi_a R_a  \label{eq:gov-Rm}
\end{equation}
\begin{equation}
C_{vm} = \xi_s C_{vs} + \xi_g C_{vg} + \xi_a C_{va} \label{eq:gov-Cvm}
\end{equation}
\begin{equation}
\xi_a = 1 - \xi \label{eq:gov-na}
\end{equation}
\begin{equation}
\xi_g = \xi \cdot \xi_{g0} \label{eq:gov-ng}
\end{equation}
\begin{equation}
\xi_s = \xi - \xi_g \label{eq:gov-ns}
\end{equation}
Where, $C_v$ is specific heat with constant volume, $R$ is gas constant. $\xi$ with subscript is mass fraction of corresponding constituent. The subscript 
$m$ represents mixture of ejected material and air, $s$ represents solid portion in the ejected material, $g$ represents gas portion in the ejected material and $a$ represents air. $\xi_{g0}$ is the mass fraction of vapor in erupted material.

>>>>

While SK-3D focuses on accurately capturing the entrainment caused by turbulent mixture with higher resolution and numerical method of higher order, PDAC takes the disequilibrium between different phases into account and hence is a true multiphase model. Another 3D model, ATHAM (Active Tracer High-Resolution Atmospheric Model) \citep{oberhuber1998volcanic} focuses more on microphysics within the plume. As pyroclastic flow is not the initial concern of ATHAM, immediate dynamic and thermodynamical equilibrium is assumed in ATHAM. The dynamic core of ATHAM solves the compressible Euler equations for momentum, pressure and temperature of the gas particle mixture \citep{oberhuber1998volcanic}. The subgrid-scale turbulence closure scheme differentiates between the horizontal and vertical directions \citep{herzog2003prognostic} to capture turbulent mixing. The cloud microphysics predicts the mass of hydrometeors in liquid and ice phase \citep{herzog1998effect}. Additional modules include radiation \citep{langmann1997radiative}, gas phase chemistry \citep{trentmann2002simulation} and gas scavenging by hydrometeors \citep{textor2003injection} were added lately. A further extension was made to include particle aggregation \citep{textor2006volcanic1, textor2006volcanic2}. However, the resolution of ATHAM is still coarse compared with SK-3D and PDAC.

@@

As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consisting of solid with different size and mixture of gases is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamical equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases does not show up under this assumption). Immediate dynamic equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term does not show up with this assumption).

@@

\subsection{Governing equations}
Based on above assumptions, the governing equations of our model are given as (which is the same as the governing equations of SK-3D \citep{suzuki2005numerical}):
\begin{align}
\dfrac{\partial \rho}{\partial t} + \nabla \cdot \left(\rho \textbf{v}\right) = 0 \label{eq:gov-cs-rho} \\
\dfrac{\partial \rho \xi}{\partial t} + \nabla \cdot \left(\rho \xi \textbf{v}\right) = 0 \label{eq:gov-cs-ks}\\
\dfrac{\partial \rho \textbf{v}}{\partial t} + \nabla \cdot \left(\rho \textbf{v} \textbf{v} + p\textbf{I}\right) = \rho \textbf{g} \label{eq:gov-cs-v} \\
\dfrac{\partial \rho E}{\partial t} + \nabla \cdot \left[\left(\rho E + p \right)\textbf{v}\right] = \rho \textbf{g} \cdot\textbf{v} \label{eq:gov-cs-e}
\end{align}
Where $\rho$ is the density, $\textbf{v}$ is the velocity, $\xi$ is the mass fraction of ejected material, $\textbf{g}$ is the gravitational acceleration, $\textbf{I}$ is a unit tensor.
$E = e + K $ is the total energy which is a summation of kinetic energy $K$ and internal energy $e$.
An additional equation is required to close the system. In this model, the equation for closing the system is the following EOS (equation of state).
\begin{equation}
p = \left(\gamma_m - 1\right)\rho e \label{eq:EOS}
\end{equation}
Where 
\begin{equation}
\gamma_m = R_m/C_{vm} + 1 \label{eq:gov-gm}
\end{equation}
\begin{equation}
R_m = \xi_g R_g + \xi_a R_a  \label{eq:gov-Rm}
\end{equation}
\begin{equation}
C_{vm} = \xi_s C_{vs} + \xi_g C_{vg} + \xi_a C_{va} \label{eq:gov-Cvm}
\end{equation}
\begin{equation}
\xi_a = 1 - \xi \label{eq:gov-na}
\end{equation}
\begin{equation}
\xi_g = \xi \cdot \xi_{g0} \label{eq:gov-ng}
\end{equation}
\begin{equation}
\xi_s = \xi - \xi_g \label{eq:gov-ns}
\end{equation}
Where, $C_v$ is specific heat with constant volume, $R$ is gas constant. $\xi$ with subscript is mass fraction of corresponding constituent. The subscript 
$m$ represents mixture of ejected material and air, $s$ represents solid portion in the ejected material, $g$ represents gas portion in the ejected material and $a$ represents air. $\xi_{g0}$ is the mass fraction of vapor in erupted material.
